GEE models

### 2017-10-15 doc/manuscript.Rmd
expanded on pls multivariable vs gee

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design an
auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE
models, though other matrices (eg. exchangeable) had similar fit (data not
shown). GEE is well suited to longitudinal cohort studies given it's capacity to
handle any missed visits and for the correlation of variables between time points. 
See the ESM Methods for more detail on GEE.
The predictor variables and continuous covariates were
scaled (mean centered and standardised). The NEFA variables were classified as 
*time-independent* (held constant) as they were measured only at the baseline 
visit, while the outcome variables were set as *time-dependent*. No imputation
was conducted on missing values and incident diabetes cases were excluded from 
the analysis.

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a;@Textor2011a] recommendations (see `r cite_sf("dag_isi")` and 
`r cite_sf("dag_issi2")`), and quasi-likelihood information criterion (QIC) [@Pan2001a]. See
the ESM Methods for more detail on QIC and the directed acyclic graphs. 
`r cite_st('qic')` shows the covariates compared using QIC. We selected the
model (M9) that had the fewest covariates and that had similar fit between the
outcome measures. The final GEE model included the covariates years from baseline
(time), sex (female or male), ethnicity (European ancestry or non-European 
ancestry), baseline age, WC, ALT, and MET (physical activity). Sex, ethnicity, 
and baseline age were time-*independent* variables. See the ESM Methods for more
detail on the covariate inclusion. After scaling,
and log-transforming the model variables, then exponentiating the GEE estimates,
the interpretation of the GEE results is as an expected percent difference in 
the outcome variable for every standard deviation (SD) increase in the predictor
variable given the covariates are held constant. With the main effect 
associations (not with a time interaction), this interpretation is applied 
consistently at each time point and thus represents associations of baseline 
NEFA for each future value of the outcomes. Since TAG is a risk factor for
diabetes and since NEFA contribute to TAG production [@Barrows2006a], TAG may
act as a mediator between NEFA and the outcomes. To determine the role of TAG in
the association between NEFA and the outcomes, TAG was included in the GEE model
in a sensitivity analysis. Lastly for the GEE models, we tested for an 
interaction with sex, ethnicity, or time by the predictor term for each outcome 
variable.

While GEE accounts for the longitudinal design of the data, this approach is
limited in that it cannot analyse the inherent multivariate nature of the
composition of the NEFA fraction, as each model only contains data from a single
fatty acid. Therefore, to confirm the results of the GEE analyses in a
multivariate environment (all NEFA variables in a single model), partial least
squares (PLS) was used with the outcomes as the dependent variable.

>>>>

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design an
auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE
models, though other matrices (eg. exchangeable) had similar fit (data not
shown). GEE is well suited to longitudinal cohort studies given it's capacity to
handle any missed visits and for the correlation of variables between time points. 
See the ESM Methods for more detail on GEE.
The predictor variables and continuous covariates were
scaled (mean centered and standardised). The NEFA variables were classified as 
*time-independent* (held constant) as they were measured only at the baseline 
visit, while the outcome variables were set as *time-dependent*. No imputation
was conducted on missing values and incident diabetes cases were excluded from 
the analysis. A GEE model is applied for each combination of outcome and NEFA (as
either mol% or nmol/ml).

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a;@Textor2011a] recommendations (see `r cite_sf("dag_isi")` and 
`r cite_sf("dag_issi2")`), and quasi-likelihood information criterion (QIC) [@Pan2001a]. See
the ESM Methods for more detail on QIC and the directed acyclic graphs. 
`r cite_st('qic')` shows the covariates compared using QIC. We selected the
model (M9) that had the fewest covariates and that had similar fit between the
outcome measures. The final GEE model included the covariates years from baseline
(time), sex (female or male), ethnicity (European ancestry or non-European 
ancestry), baseline age, WC, ALT, and MET (physical activity). Sex, ethnicity, 
and baseline age were time-*independent* variables. See the ESM Methods for more
detail on the covariate inclusion. After scaling,
and log-transforming the model variables, then exponentiating the GEE estimates,
the interpretation of the GEE results is as an expected percent difference in 
the outcome variable for every standard deviation (SD) increase in the predictor
variable, for the same covariate values (e.g. same age or sex). With the main effect 
associations (not with a time interaction), this interpretation is applied 
consistently at each time point and thus represents associations of baseline 
NEFA for each future value of the outcomes. Since TAG is a risk factor for
diabetes and since NEFA contribute to TAG production [@Barrows2006a], TAG may
act as a mediator between NEFA and the outcomes. To determine the role of TAG in
the association between NEFA and the outcomes, TAG was included in the GEE model
in a sensitivity analysis. Lastly for the GEE models, we tested for an 
interaction with sex, ethnicity, or time by the predictor term for each outcome 
variable.

While GEE accounts for the longitudinal design of the data, this approach is
limited in that it cannot analyse the inherent multivariate nature of the
composition of the NEFA fraction, as each model only contains data from a single
fatty acid. Therefore, to confirm the results of the GEE analyses in a
multivariate environment (with all NEFA variables in a single model), partial least
squares (PLS) was used with the outcomes as the dependent variable.
